# Functions {#functions}

This chapter contains the basics of how to write a function, links to more complete instructions, and examples of functions.

```{r, include=FALSE}
# Every file must contain at least one R chunk due to the linting process.
```

## User defined functions (UDFs)

A non-exhaustive list of possible reasons for writing your own UDF,

* you are repeating some code chunk changing only a few variables each time.
* you want to apply the same set of commands to multiple instances of some object: it might be a dataframe, a text string, a document, an image etc.
* tidier code, you can write the functions in a separate R file and load these functions by running that file when you load other packages.
* easier to update, just update the function rather than edit potentially multiple code chunks.
* prevents copy/paste errors
* you are writing a package/library

## Top tips for writing functions

* name the function something sensible that gives you a clue what it does, usually containing a verb, and making sure that it doesn't conflict with names of other functions in use.
* write some description of what the function is intended for and the expected outcome, including what it needs to work properly, eg it might take an integer value argument only, and therefore fail if given a double.
* make your function as independent as possible, use `name_space::function` syntax if it uses functions from other packages so they don't have to be loaded (they will have to be installed though).
* recommend passing named arguments rather than relying on global variables (again name your arguments clearly).
* recommend giving arguments default values, this shows an example of the type of variable that is expected here.
* note that variables assigned inside a function will not appear in the global environment unless explicitly returned.
* using an explicit return statement is useful if the function is complex, otherwise the output will be the final evaluated expression.

## Function structure

Below is an example of the syntax for defining a function named `state_animal_age`. The list of named arguments are included in brackets `()` with defaults. The function code sits inside braces `{}`, and returns a single string of the concatenated inputs. 

```{r, echo=TRUE}
state_animal_age <- function(animal = "animal type", age = 0) {
  #function takes animal name as a string and current age and pastes them together
  #output is single string
  return( paste(animal, age, sep = ': ') )
}

#run function with defaults
state_animal_age()

#run function passing inputs to argument names, order doesn't matter
state_animal_age(age = 2, animal = "potoo")

#run function passing inputs by position, order matters
state_animal_age("tortoise", 169)
```

## Function with `purrr` structure

The scenario in which I most find myself writing functions is when I want to do the same thing to a set of similar objects. I write a function that does what I want to one object and then use `purrr::map` (or one of its siblings) to iterate through the complete list of objects. This greatly simplifies the entire process. For examples see Chapter 3 **.xlsx and .xls** section and below.

Using the function above that takes two inputs I can use `purrr::map2` to iterate over two lists of these input values. Note that the corresponding values in each list form a pair of inputs for the function, so we get three outputs. Consider how this is different to iterating over nested lists, where, in this case, we would get nine outputs.  

```{r, echo=TRUE}
animal_list = c('cat', 'dog', 'elephant')
age_list = c(12, 7, 42)

purrr::map2(.x = animal_list, .y = age_list, .f = state_animal_age)
```

## Function to produce multiple graphs

A generic function to create multiple graphs based on a list e.g. a graph for each local authority.

The function creates a list to iterate over to produce multiple graphs, in this case a list of local authorities.

The function then uses a loop to basically create a graph for each value in the list. 

The graphs are created using ggplot, where the data being used to create the graphs is subset based on the value it is currently using from the list. So, in this case the data is subset based on each local authority in the list and a graph is created using this subset of data. This is then looped for each local authority in the list.

The function saves each graph as a png file in the working directory.


```{r, eval = FALSE, echo = TRUE}

# create graphing function
    multiple.graph <- function(df, na.rm = TRUE, ...){
      
      # create list of local authorities in data to loop over 
      la_list <- unique(df$local_authority)
      
      # create for loop to produce ggplot2 graphs 
      for (i in seq_along(la_list)) { 
        
        # create plot for each LA in df 
        plot <- 
          ggplot(subset(df, df$local_authority==la_list[i]),
                 aes(x = year)) +
          geom_line(aes(y = serious_accidents), colour = "blue", size = 2) +
          ylab("serious accident count") +
          xlab("Year") +
          theme_bw() +
          ggtitle(paste(la_list[i], "comparison chart"))    # this produces a dynamic title which is updated based on the local authority currently being used from the list

        
         # save plots as .png
        ggsave(plot, file=paste(la_list[i], ".png", sep=''), scale=2)
        
         # print plots to screen
    print(plot)
  }
    }

## run function
multiple.graph(dataframe3)

```


## Further reading

[Functions chapter of R for Data Science by Hadley Wickham](https://r4ds.had.co.nz/functions.html)

## Function to create folder structure for R project

